No proof nets for MLL with units\\[5pt]\Large Proof equivalence in MLL is \textsc{pspace}-complete

### 2014-01-19 main/LICS.tex
working on "soundness"

\begin{enumerate}
	\item
sub-sequents are strictly nested: if $\named w\parr$ occurs in $s(v)$ then $s(w)<s(v)$; and
	\item
for each $\named v\parr$ the graph $\sigma(v)=(s(v),\links|_{s(v)})$ becomes a tree when all immediate sub-sequents are contracted.
\end{enumerate}

@@

Finally, the standard notion of \emph{balance} is useful in constructing proof nets without $\parr$-connectives

@@

An unbalanced sequent is uninhabited: a positive balance guarantees a cycle in any switching graph, for any linking, while a negative balance similarly guarantees disconnectedness.
%
%An early conjecture of Girard, which turned out to be false, was that a sequent is inhabited if and only if it is balanced. [[FIND CITATION (probably TCS87)]]
%

@@

%
\emph{Non-deterministic constraint graph reconfiguration} or \emph{\textsc{ncg}-reconfiguration} is the problem of deciding when two configurations for a constraint graph are connected in the graph of total configurations and reconfiguration steps.
%
\emph{Partial \textsc{ncg}-reconfiguration} is the same problem for partial configurations.

@@

By Proposition~\ref{prop:partial simulates total reconfiguration} the following is then immediate.

\begin{proposition}
Partial \textsc{ncg}-reconfiguration is \textsc{pspace}-complete.
%\textsc{pncg}-reconfiguration is \textsc{pspace}-complete.
\end{proposition}

We will demonstrate the \textsc{pspace}-hardness of MLL proof equivalence by an encoding of partial \textsc{ncg}-reconfiguration in MLL proof nets.

@@

\begin{definition}
\label{def:graph encoding}
The \emph{encoding} $\itn G$ of a constraint graph $G=(V,E,c,v,w)$ is a sequent constructed as follows.
%
Let $V=\{v_1,\dotsc,v_n\}$ and $E=\{e_1,\dotsc,e_m\}$ where $|V|=n$ and $|E|=m$.

@@

	\item
To exchange a weight absorber $w$ linked from a $\named d\bot$ outside $\itn e$ for the target $v$ of $e_1$, and attach $e_0$ to $w$, attach $d$ to $v$.
%
At this point, $e_1$ forms the only connect between $\itn e$ and the remainder of the proof net; to reinstate $e_0$ in this role, do as above, producing a net effect of cycling the targets of $e_1$, $e_2$, and $d$.

	\item
Finally, to permute weight absorbers $v$ and $w$ linked by $\itn e$, attach $e_0$ to the indicator of a vertex-gadget where also another edge-gadget $\itn d$ is connected, with indicator vertex $d_0$ and an arbitrary other vertex $d_1$.
%
Connect $d_0$ to $v$, and apply Lemma~\ref{lem:double exchange} to permute $v$ and $w$, as well as exchanging the targets of $d_0$ and $d_1$.
%
A second such exchange is needed to re-attach $d_0$ and $d_1$ to their original targets.

@@

Let $\itn\gamma'$ be $\itn\gamma$ where the first two weight absorbers are exchanged.

\begin{lemma}
If $\gamma\sim\delta$ for total configurations $\gamma$ and $\delta$, then either $\itn\gamma\sim\itn\delta$ or $\itn\gamma\sim\itn\delta'$.
\end{lemma}

\begin{IEEEproof}
We will prove the case where $\gamma\sim\delta$ is a single reconfiguration step; the general case follows because the proof goes through also when $\itn\gamma'$ replaces $\itn\gamma$ in the statement of the lemma.
%
Let $\gamma$ and $\delta$ agree on every edge except $e$, where $\gamma(e)=v$ and $\delta(e)=w$.
%
Firstly, using Lemma~\ref{lem:octopus roll}, for the edges $d$ other than $e$ such that $\gamma(d)=v$, the weight elements of the edge-gadgets $\itn d$ may be linked to the constraint elements of $\itn v$, in accordance with the target configuration $\itn\delta$.
%
Since $e$ is mobile in $\gamma$, the weights of the edges $d$ suffice to fill the inflow constraint of $v$, and correspondingly the weight elements of edge-gadgets $\itn d$ suffice to fill the constraint elements of $\itn v$, so that $\itn e$ is free.
%
Next, the indicator vertex of $\itn e$, which links to the indicator of $\itn v$, is re-attached to the indicator of $\itn w$.
%
Again using Lemma~\ref{lem:octopus roll}, the weight elements of edge-gadgets connected to $\itn w$, including $\itn e$, may be linked in accordance with $\itn\delta$.
%
The resulting proof net is $\itn\delta$ modulo a permutation of weight absorbers; then it is equivalent to either $\itn\delta$ or $\itn\delta'$ by Lemma~\ref{lem:permute weight absorbers}.
\end{IEEEproof}

@@

@@

\begin{IEEEproof}
Since vertex-gadgets are joined by a tensor, the lemma is immediate from \cite[Proposition 1]{Bellin-vandeWiele-1995}.
\end{IEEEproof}

\begin{lemma}
\label{lem:appropriate edge weights}
In a proof net for $\itn G$, for each vertex $v$, the weights of the appropriate edge-gadgets in the empire of $\itn v$ are equal to or greater than the constraint of $v$.
\end{lemma}

\begin{IEEEproof}
Let $|V|=n$ and $|E|=m$, and consider the vertex $v_i$ and an edge $e$ connecting vertices $v_a$ and $v_b$ where $i\neq a,b$.
%
Each constraint element in $\itn{v_i}$ is an instance of the formula
\[
	C = \bigparr\big(1^{3i+2}\big) \tn \bigparr\big(1^{3(n-i)+3}\big)~.
\]
The two $\parr$-subformulae have a balance of $-3i-1$ and $-3(n-i)-2$ respectively.
%
Each weight element in $\itn{e}$ is an instance of 
\[
	W = \bigtn\big(\bot^{3a+2}\big)\parr\bigtn\big(\bot^{3(b-a)+1}\big)\parr\bigtn\big(\bot^{3(n-b)+3}\big)~.
\]
The three $\tn$-subformulae have a net balance of $3a+1$, $3(b-a)$, and $3(n-b)+2$ respectively.
%
In pairs, they have a net balance of $3b+1$ (1st and 2nd subformula), $3(a+n-b)+3$ (1st and 3rd), and $3(n-a)+2$ (2nd and 3rd).
%
Since $i\neq a,b$, and since no $\tn$-subformula of $W$ can connect to more than one $\parr$-subformula of $C$, it follows that $W$ can balance the scope of at most one of both subformulae of $C$.

@@

\begin{lemma}
If $\links\sim\links'$ are proof nets for $\itn G$ then $\coitn\links\sim\coitn{\links'}$.
\end{lemma}

\begin{IEEEproof}
A single permutation $\links\sim\links'$ on proof nets may move a number of edge-gadgets $\itn{e_1},\dotso,\itn{e_n}$ in three ways: 1) out of the empire of a vertex-gadget $\int v$, 2) into the empire of a vertex-gadget $\int w$, or 3) both.
%
Then in both $\coitn\links$ and $\coitn{\links'}$ the edges $e_1$ through $e_n$ are mobile, since by Lemma~\ref{lem:appropriate edge weights} the empires of the vertex-gadgets $\itn v$ (in cases 1 and 3) and $\itn w$ (in cases 2 and 3) contain appropriate edge-gadgets other than $\itn{e_1}$ through $\itn{e_n}$ of sufficient combined weight.
%
It follows that in the graph $G$ the edges $e_1$ through $e_n$ can be moved away from $v$ and/or onto $w$ one at a time.

>>>>

\begin{enumerate}
	\item
sub-sequents either disjoint or strictly nested: if $s(v)\cap s(w)\neq\varnothing$ then $s(w)<s(v)$ or $s(v)<s(w)$; and
	\item
for each $\named v\parr$ the graph $\sigma(v)=(s(v),\links|_{s(v)})$ becomes a tree when all immediate sub-sequents are contracted.
\end{enumerate}

@@

Finally, the standard notion of \emph{balance} provides an easy way of identifying some unprovable sequents.

@@

\begin{proposition}
\label{prop:unbalanced then uninhabited}
An unbalanced sequent is uninhabited.
\end{proposition}

%A positive balance guarantees a cycle in any switching graph, for any linking, while a negative balance similarly guarantees disconnectedness.
%
%An early conjecture of Girard, which turned out to be false, was that a sequent is inhabited if and only if it is balanced. [[FIND CITATION (probably TCS87)]]
%

@@

\emph{Non-deterministic constraint graph reconfiguration} or \emph{\textsc{ncg}-reconfiguration} is the problem of deciding whether two total configurations for a constraint graph are connected by a sequence of reconfiguration steps.
%
%By the above proposition, \textsc{ncg}-reconfiguration is unaffected by inclusion or exclusion of partial configurations.

@@

%By Proposition~\ref{prop:partial simulates total reconfiguration} the following is then immediate.
%
%\begin{proposition}
%Partial \textsc{ncg}-reconfiguration is \textsc{pspace}-complete.
%%\textsc{pncg}-reconfiguration is \textsc{pspace}-complete.
%\end{proposition}

We will demonstrate the \textsc{pspace}-hardness of MLL proof equivalence by an encoding of \textsc{ncg}-reconfiguration in MLL proof nets.

@@

\begin{definition}
\label{def:graph encoding}
The \emph{encoding} $\itn G$ of a constraint graph $G$ is a sequent constructed as follows.
%
Let $G=(V,E,c,v,w)$, and let $V=\{v_1,\dotsc,v_n\}$ and $E=\{e_1,\dotsc,e_m\}$ where $|V|=n$ and $|E|=m$.

@@

	\item
To exchange a weight absorber $w$ linked from a $\named d\bot$ outside $\itn e$ for the target $v$ of $e_1$, first attach $e_0$ to $w$, then re-attach $d$ to $v$.
%
At this point, $e_1$ forms the only connect between $\itn e$ and the remainder of the proof net; to reinstate $e_0$ in this role, do as above, producing a net effect of cycling the targets of $e_1$, $e_2$, and $d$.

	\item
Finally, to permute weight absorbers $v$ and $w$ linked by $\itn e$, attach $e_0$ to the indicator target of a vertex-gadget where also another edge-gadget $\itn d$ is connected, with indicator $d_0$ and an arbitrary other $\bot$-occurrence $d_1$.
%
Connect $d_0$ to $v$, and apply Lemma~\ref{lem:double exchange} to permute $v$ and $w$, as well as exchanging the targets of $d_0$ and $d_1$.
%
A second such exchange is needed to re-attach $d_0$ and $d_1$ to their original targets.

@@

For any configuration $\gamma$ let $\itn\gamma'$ denote $\itn\gamma$ where the first two weight absorbers are exchanged.

\begin{lemma}
\label{lem:completeness}
If $\gamma\perm\delta$ for total configurations $\gamma$ and $\delta$, then either $\itn\gamma\perm\itn\delta$ or $\itn\gamma\perm\itn\delta'$.
\end{lemma}

\begin{IEEEproof}
By Proposition~\ref{prop:partial simulates total reconfiguration} we may assume that $\gamma\perm\delta$ by a sequence of reconfiguration steps over total configurations.
%
We will prove that if $\gamma\perm*\delta$ then $\itn\gamma\perm\itn\delta$ or $\itn\gamma\perm\itn\delta'$.
%
The same proof will show the corresponding case for $\itn\gamma'$ instead of $\itn\gamma$, so that the general case for $\gamma\perm\delta$ follows by transitivity.

Let $\gamma$ and $\delta$ agree on every edge except $e$, where $\gamma(e)=v$ and $\delta(e)=w$.
%
Firstly, using Lemma~\ref{lem:octopus roll}, for the edges $d$ other than $e$ such that $\gamma(d)=v$, the weight elements of the edge-gadgets $\itn d$ may be linked to the constraint elements of $\itn v$, in accordance with the target configuration $\itn\delta$.
%
Since $e$ is mobile in $\gamma$, the weights of the edges $d$ suffice to fill the inflow constraint of $v$, and correspondingly the weight elements of edge-gadgets $\itn d$ suffice to fill the constraint elements of $\itn v$, so that $\itn e$ is free.
%
Next, the indicator vertex of $\itn e$, which links to the indicator target of $\itn v$, is re-attached to the indicator target of $\itn w$.
%
Again using Lemma~\ref{lem:octopus roll}, the weight elements of edge-gadgets connected to $\itn w$, including $\itn e$, may be linked in accordance with $\itn\delta$.
%
The resulting proof net is $\itn\delta$ modulo a permutation of weight absorbers; then it is equivalent to either $\itn\delta$ or $\itn\delta'$ by Lemma~\ref{lem:permute weight absorbers}.
\end{IEEEproof}

@@

It will be shown that proof net rewiring $\itn\gamma\perm\itn\delta$ is sound for \textsc{ncg}-reconfiguration, $\gamma\perm\delta$.
%
To each proof net $\links$ for an encoded constraint graph $\itn G$ we will associate a partial configuration $\gamma=\coitn\links$, such that 1) a rewiring step $\links\perm*\links$ corresponds to a reconfiguration path $\coitn\links\perm\coitn{\links'}$, and 2) it is a retraction of the encoding of configurations, $\coitn{\itn\gamma}=\gamma$.

The configuration $\coitn\links$ will assign an edge $e$ to a vertex $v$ when, in the proof net $(\links,\itn G)$, the edge-gadget $\itn e$ is in the empire of the vertex-gadget $\itn v$.
%
Firstly, it will be shown that $\coitn\links$ is a partial function.

\begin{proposition}[{\cite[Proposition 1.i]{Bellin-vandeWiele-1995}}]
\label{prop:tensor has disjoint subnets}
In a proof net, if vertices $v$ and $w$ are joined by a tensor, then any two subnets of which they are respective ports are disjoint.
\end{proposition}

@@

\begin{IEEEproof}
Since vertex-gadgets are joined by a tensor, the lemma is immediate from Proposition~\ref{prop:tensor has disjoint subnets}.
\end{IEEEproof}

Next, it will be shown that the appropriate edge-gadgets in the empire of a vertex-gadget $\itn v$ contain sufficient weight elements to fill the constraint elements of $\itn v$.

\begin{lemma}
\label{lem:appropriate edge weights}
In a proof net for $\itn G$, for each node $v$ in $G$, the weights of the appropriate edge-gadgets in the empire of $\itn v$ are equal to or greater than the constraint of $v$.
\end{lemma}

\begin{IEEEproof}
Let $|V|=n$ and $|E|=m$, and consider the vertex $v_i$ and an edge $e$ connecting vertices $v_a$ and $v_b$ where $i\neq a,b$.
%
Each constraint element in $\itn{v_i}$ is an instance of the formula
\[
	C = \bigparr\big(1^{3i+2}\big) \tn \bigparr\big(1^{3(n-i)+3}\big)~.
\]
The two $\parr$-subformulae have a balance of $-3i-1$ and respectively $-3(n-i)-2$.
%
Each weight element in $\itn{e}$ is an instance of 
\[
	W = \bigtn\big(\bot^{3a+2}\big)\parr\bigtn\big(\bot^{3(b-a)+1}\big)\parr\bigtn\big(\bot^{3(n-b)+3}\big)~.
\]
For each of the three $\tn$-subformulae, adding it to the scope of a $\parr$ will change its balance by $3a+1$, $3(b-a)$, and $3(n-b)+2$ respectively.
%
In pairs, they add to the balance of a scope by $3b+1$ (1st and 2nd subformula), $3(a+n-b)+3$ (1st and 3rd), and $3(n-a)+2$ (2nd and 3rd).
%
Since $i\neq a,b$, and since no $\tn$-subformula of $W$ can connect to more than one $\parr$-subformula of $C$ (by Proposition~\ref{prop:tensor has disjoint subnets}), it follows that $W$ can balance the scope of at most one of both $\parr$-subformulae of $C$.

@@

\begin{lemma}
If $\links\perm\links'$ are proof nets for $\itn G$ then $\coitn\links\perm\coitn{\links'}$.
\end{lemma}

\begin{IEEEproof}
The proof will consider the case where $\links\perm*\links'$; the general case follows by transitivity.
%
A single permutation $\links\sim\links'$ on proof nets may move a number of edge-gadgets $\itn{e_1},\dotso,\itn{e_n}$ in three ways: 1) out of the empire of a vertex-gadget $\int v$, 2) into the empire of a vertex-gadget $\int w$, or 3) both.
%
Then in both $\coitn\links$ and $\coitn{\links'}$ the edges $e_1$ through $e_n$ are mobile, since by Lemma~\ref{lem:appropriate edge weights} the empires of the vertex-gadgets $\itn v$ (in cases 1 and 3) and $\itn w$ (in cases 2 and 3) contain appropriate edge-gadgets other than $\itn{e_1}$ through $\itn{e_n}$ of sufficient combined weight.
%
It follows that in the graph $G$ the edges $e_1$ through $e_n$ can be moved away from $v$ and/or onto $w$ one at a time.
